### 2010-10-12 Thesis/Chapter5.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{The total Lagrangian explicit dynamics (TLED) algorithm}
\label{chap5}
\begin{shortAbstract}
In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. Their work constituted a step towards the simulation of entire organs in real-time. Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method. However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation. \cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind. We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well. This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}. This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions. Their GPU implementations will be the object of the next chapter. 
\end{shortAbstract}

@@

\noindent The required stress $ \mathbf{S} $ may be obtained via differentiation with respect to strain:
\begin{equation}
\label{chap5:stress}
\mathbf{S} = 2 \pd{\hat{\Psi}(\Psi , t)}{\mathbf{C}} = \int_0^t \alpha (t-s) \left( 2 \pd{}{s} \pd{\Psi}{\mathbf{C}} \right) ds = \int_0^t \alpha (t-s) \pd{\boldsymbol \Phi}{s} ds,
\end{equation}
where $ \mathbf{C} $ is the right Cauchy-Green deformation tensor, and we have introduced $ \boldsymbol \Phi \, {\buildrel\rm	def\over=} \, 2 \partial \Psi / \partial \mathbf{C} $ as the instantaneous hyperelastic stress response.

@@

	\subsubsection*{Transversely isotropic materials}
Transversely isotropic materials are characterised by a single preferred direction $\mathbf{a}_0$ in the reference configuration; the mechanical response is isotropic in the plane orthogonal to this direction. The strain energy is then a function of both $  \mathbf{C} $ and a structure tensor $\mathbf{A}_0 \, {\buildrel\rm	def\over=} \,  \mathbf{a}_0 \otimes \mathbf{a}_0$, where $\otimes$ denotes a tensor product. Analogous to the isotropic case, we consider strain energy functions of the form	
\begin{equation}
\label{chap5:TIstrainEnergy}
\Psi (\mathbf{C}, \mathbf{A}_0) = \Psi^{iso}(\bar{I}_1, \bar{I}_2, \bar{I}_4, \bar{I}_5) + \Psi^{vol}(J),
\end{equation}
where $ \bar{I}_4 = \mathbf{a}_0 \cdot \mathbf{\bar{C}} \mathbf{a}_0$ and $ \bar{I}_5 = \mathbf{a}_0 \cdot \mathbf{\bar{C}}^2 \mathbf{a}_0 $ are pseudo-invariants of $ \mathbf{\bar{C}} $ and $ \mathbf{A}_0 $.

>>>>

\chapter{The total Lagrangian explicit dynamics algorithm}
\label{chap5}
\begin{shortAbstract}
In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation. The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation. Their work constituted a step towards the simulation of entire organs in real-time. Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method. However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation. \cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm. With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind. We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well. This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}. This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions. Their GPU implementations will be the object of the next chapter. 
\end{shortAbstract}

@@

\noindent The required stress $ \mathbf{S} $ may be obtained via differentiation with respect to strain:
\begin{equation}
\label{chap5:stress}
\mathbf{S} = 2 \pd{\hat{\Psi}(\Psi , t)}{\mathbf{C}} = \int_0^t \alpha (t-s) \left( 2 \pd{}{s} \pd{\Psi}{\mathbf{C}} \right) ds = \int_0^t \alpha (t-s) \pd{\boldsymbol \Phi}{s} ds,
\end{equation}
where $ \mathbf{C} $ is the right Cauchy-Green deformation tensor, and we have introduced $ \boldsymbol \Phi \, {\buildrel\mathrm	def\over=} \, 2 \partial \Psi / \partial \mathbf{C} $ as the instantaneous hyperelastic stress response.

@@

	\subsubsection*{Transversely isotropic materials}
Transversely isotropic materials are characterised by a single preferred direction $\mathbf{a}_0$ in the reference configuration; the mechanical response is isotropic in the plane orthogonal to this direction. The strain energy is then a function of both $  \mathbf{C} $ and a structure tensor $\mathbf{A}_0 \, {\buildrel\mathrm	def\over=} \,  \mathbf{a}_0 \otimes \mathbf{a}_0$, where $\otimes$ denotes a tensor product. Analogous to the isotropic case, we consider strain energy functions of the form	
\begin{equation}
\label{chap5:TIstrainEnergy}
\Psi (\mathbf{C}, \mathbf{A}_0) = \Psi^{iso}(\bar{I}_1, \bar{I}_2, \bar{I}_4, \bar{I}_5) + \Psi^{vol}(J),
\end{equation}
where $ \bar{I}_4 = \mathbf{a}_0 \cdot \mathbf{\bar{C}} \mathbf{a}_0$ and $ \bar{I}_5 = \mathbf{a}_0 \cdot \mathbf{\bar{C}}^2 \mathbf{a}_0 $ are pseudo-invariants of $ \mathbf{\bar{C}} $ and $ \mathbf{A}_0 $.

### 2010-10-12 Thesis/Chapter6.tex
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Compression of a cube}
The second example involved compression of a cube of edge length $0.1\,$m (comparable to some human organs). The cube was imagined to be fixed to opposing load platens and free on the remaining four faces. Compression was applied along the $x$-axis. A stress-relaxation type test protocol was simulated, in which the cube was compressed by $30\, \%$ in 0.5 seconds and the compression was held for a further 4.5 seconds. We demonstrate the effects of anisotropy on the deformation response by comparing the deformed shapes of anisotropic models with $\mathbf{a}_0 = [0\, 1\, 0] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^y $ and $\mathbf{a}_0 = [0\, \frac{1}{\sqrt{2}} \, \frac{1}{\sqrt{2}}] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^{yz} $  with that of an isotropic model (with $\eta = 0$).

@@

		\subsubsection*{General GPU performance}
Next, we assessed the computational performance of the GPU implementation and efficiency of the constitutive update scheme by measuring computation times for a range of mesh densities. Using the cube geometry as above we generated models with mesh sizes ranging from $3\,993$ DOF ($1\,331$ nodes, $1\,000$ hexahedral elements) to $177\,957$ DOF ($59\,319$ nodes, $54\,872$ hexahedral elements). We compared the GPU solution times for a single time step with those of a CPU-based implementation developed using C++. The test machine included an Intel Core2Duo 2.4GHz CPU, 2GB RAM, and an NVIDIA GeForce 8800GTX GPU. Three constitutive models were considered: the transversely isotropic viscoelastic model (TIV), TIV minus the viscoelastic terms (TIE), and TIE minus the anisotropy terms (NHE). Comparison of times for TIV and TIE indicates the computational load introduced by the developed viscoelastic constitutive update scheme. Comparison of times for TIE and NHE indicates the computational load introduced by material anisotropy. Figure~\ref{chap6:fig-timingsGPU} shows the results of the experiments. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13.5cm]{chapter6/timingsGPU.pdf}
\end{center}
\caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE. (b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.} 
\label{chap6:fig-timingsGPU}
\end{figure}

>>>>

	\subsection{Compression of a cube}
The second example involved compression of a cube of edge length $0.1\,$m (comparable to some human organs). The cube was imagined to be fixed to opposing load platens and free on the remaining four faces. Compression was applied along the $x$-axis. A stress-relaxation type test protocol was simulated, in which the cube was compressed by $30\, \%$ in 0.5 seconds and the compression was held for a further 4.5 seconds. We demonstrate the effects of anisotropy on the deformation response by comparing the deformed shapes of anisotropic models with $\mathbf{a}_0 = [0\, 1\, 0] \, {\buildrel\mathrm	def\over=} \, \mathbf{a}_0^y $ and $\mathbf{a}_0 = [0\, \frac{1}{\sqrt{2}} \, \frac{1}{\sqrt{2}}] \, {\buildrel\mathrm	def\over=} \, \mathbf{a}_0^{yz} $  with that of an isotropic model (with $\eta = 0$).

@@

		\subsubsection*{General GPU performance}
Next, we assessed the computational performance of the GPU implementation and efficiency of the constitutive update scheme by measuring computation times for a range of mesh densities. Using the cube geometry as above we generated models with mesh sizes ranging from $3\,993$ DOF ($1\,331$ nodes, $1\,000$ hexahedral elements) to $177\,957$ DOF ($59\,319$ nodes, $54\,872$ hexahedral elements). We compared the GPU solution times for a single time step with those of a CPU-based implementation developed using C++. The test machine included an Intel Core2Duo 2.4GHz CPU, 2GB RAM, and an NVIDIA GeForce 8800GTX GPU. Three constitutive models were considered: the transversely isotropic viscoelastic model (TIV), TIV minus the viscoelastic terms (TIE), and TIE minus the anisotropy terms (NHE). Comparison of times for TIV and TIE indicates the computational load introduced by the developed viscoelastic constitutive update scheme. Comparison of times for TIE and NHE indicates the computational load introduced by material anisotropy. Figure~\ref{chap6:fig-timingsGPU} shows the results of the experiments. 	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU1.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU2.pdf}} \\
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU3.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU4.pdf}} \\
\caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE. (b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.} 
\label{chap6:fig-timingsGPU}
\end{figure}
